---
title: "CN1 - Processed Environmental Data"
author: "Madeline Eppley"
date: "3/25/2024"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData")
```

### Load required packages. 
```{r}
library("dplyr") #Used for working with data frames
library("lubridate") #Used for time-date conversions
library("readr") #Used to read the CSV file
library("ggplot2") 
```

### Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  
```{r}
#Data was downloaded on 3/25/2024
#Source - https://data.novascotia.ca/Nature-and-Environment/Nova-Scotia-Water-Quality-Data-Station-Locations-M/svms-mkst
#The site was sampled continuously 

#Create text strings with metadata information that we want to include in the final data frame. 
download_date <- ("03-25-2024")
source_description <- ("Nova Scotia Gov Water Quality Data")
site_name <- ("CN1") #Use site code with site number based on lat position and state
collection_type <- ("continuous")
```

### Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names. 
```{r}
#The file we will be working with is from Salt Bay, Canada. The ID_Site for this site is CN1. 
raw_CN1_sal <- read_csv("/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/CN1-raw_sal.csv")
raw_CN1_temp <- read_csv("/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/CN1-raw_temp.csv")

# View how the data is stored. Note the variable names and the format and units that the data are stored in.  
spec(raw_CN1_sal)
#View(raw_CN1_sal)

spec(raw_CN1_temp)
#View(raw_CN1_temp)
```

### Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (1/1/1970). This makes comparisons easy and helps to standardizes values. 
```{r}

#Convert to POSIXct format. Store it into a column named datetime in the data frame.
raw_CN1_sal$datetime <- as.POSIXct(raw_CN1_sal$TIMESTAMP, "%Y/%m/%d %I:%M:%S %p", tz = "")
raw_CN1_temp$datetime <- as.POSIXct(raw_CN1_temp$TIMESTAMP, "%Y/%m/%d %I:%M:%S %p", tz = "")

#Print the new data frame and examine to make sure the new datetime column is in the correct format. 
head(raw_CN1_sal)
head(raw_CN1_temp)
```

### Analyze the ranges of all of our variables of interest - time, salinity, and temperature. Make sure that the latitude and longitude values are consistent for a static collection site. This is a quick check so we can determine how to conduct the next filtering step. 
```{r}
#Standardize column and variable names. We will use "temp" for temperature in degrees C, "salinity" for salinity in parts per thousand (ppt), "lat" for latitude in degrees, and "lon" for longitude in degrees. 
#Use the dyplr format to rename multiple columns in the format "dataframe %>% rename("new name 1" = "old name 1", "new name 2", "old name 2")
raw_CN1_sal <- raw_CN1_sal %>% rename("salinity" = "VALUE")
raw_CN1_temp <- raw_CN1_temp  %>% rename("temp" = "VALUE")

#Store variables that we will include in the final data frame
lat <- 43.79075
lon <- -65.83619
firstyear <- 2020
finalyear <- 2021
```


### Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 
```{r}
#Filter the data between the values of 0 and 40 for both salinity and temperature. 
filtered_CN1_sal <- raw_CN1_sal %>%
    filter(between(salinity, 0, 42)) 
           
filtered_CN1_temp <- raw_CN1_temp %>%
    filter(between(temp, 0, 40))

# Sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(filtered_CN1_sal$salinity))
print(summary(filtered_CN1_temp$temp))

#Store our data into a variable name with just the site name. 
CN1_temp <- filtered_CN1_temp
CN1_sal <- filtered_CN1_sal


# we have NAs in the our salinity data frame in the datetime column - need to remove these
count.nas_sal <- is.na(CN1_sal$datetime) # store our NAs in a variable
summary(count.nas_sal) # we have 12 NAs that are stored as "TRUE" in our count.nas

nrow(CN1_sal) # figure out how many rows we have in the original df: 104905
which(count.nas_sal == TRUE) # find the number of NA rows that we need to remove: 12
CN1_sal <- CN1_sal[-c(38074, 38075, 38076, 38077, 38078, 38079, 
                      90414, 90415, 90416, 90417, 90418, 90419), ] # remove the rows
nrow(CN1_sal) # check the new number of rows in the dataframe with the NAs removed
check_sal <- 104905-104893 # the value of check should be 12
check_sal # cool, we removed the 12 NA rows!

# we have NAs in the our temperature data frame in the datetime column - need to remove these
count.nas_temp <- is.na(CN1_temp$datetime) # store our NAs in a variable
summary(count.nas_temp) # we have 6 NAs that are stored as "TRUE" in our count.nas

nrow(CN1_temp) # figure out how many rows we have in the original df: 56634
which(count.nas_temp == TRUE) # find the number of NA rows that we need to remove: 6
CN1_temp <- CN1_temp[-c(38266, 38267, 38268, 38269, 38270, 38271), ] # remove the rows
nrow(CN1_temp) # check the new number of rows in the dataframe with the NAs removed
check_temp <- 56634-56628 # the value of check should be 12
check_temp # cool, we removed the 6 NA rows!


```

### Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}
salplot <- ggplot(CN1_sal, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0,40) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for CN1 - Salt Bay, Canada") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()

salplot
```


```{r temperature-plot}
tempplot <- ggplot(CN1_temp, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 45) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for CN1 - Salt Bay, Canada") +
    scale_color_manual(values = c( "Temperature (C)" = "red")) +
    theme_minimal()


tempplot
```

### We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.
```{r}
#Calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
CN1_envrmonth_sal <- CN1_sal %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity))
      
CN1_envrmonth_temp <- CN1_temp %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(      
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp))
  
print(CN1_envrmonth_sal)
print(CN1_envrmonth_temp)


#Calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
CN1_envryear_sal <- CN1_sal %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity))

CN1_envryear_temp <- CN1_temp %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp))

print(CN1_envryear_sal)
print(CN1_envryear_temp)

```
### Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot - salinity}
timeplot <- ggplot(CN1_envrmonth_sal, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Salinity Timeplot for CN1 - Salt Bay, Canada") +
    ylim(1,12) +
    theme_minimal()

timeplot
```

### Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot - temperature}
timeplot <- ggplot(CN1_envrmonth_temp, aes(x = year)) +
    geom_point(aes(y = month, color = length_temp), size = 4) +
    labs(x = "Time", y = "Month", title = "Temperature Timeplot for CN1 - Salt Bay, Canada") +
    ylim(1,12) +
    theme_minimal()

timeplot
```

### We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables. 

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set

```{r}
#Calculate temperature variables. 
temp_quantile_10 <- quantile(CN1_temp$temp, 0.1)
temp_quantile_90 <- quantile(CN1_temp$temp, 0.9)
Mean_Annual_Temperature_C <- mean(CN1_temp$temp)
Mean_max_temperature_C <- mean(CN1_envryear_temp$max_temp)
Mean_min_temperature_C <- mean(CN1_envryear_temp$min_temp)
Temperature_st_dev <- sd(CN1_temp$temp)
Temperature_n <- nrow(CN1_temp)
Temperature_years <- nrow(CN1_envryear_temp)

#Create a data frame to store the temperature results
CN1_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, Temperature_st_dev, Temperature_n, Temperature_years,temp_quantile_10, temp_quantile_90, collection_type)
print(CN1_temp)

# Write to the combined file with all sites 
write.table(CN1_temp, "/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/Seascape_Processed/seascape_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE) # The column names should be changed to FALSE after 1st row is added to the data frame

# Write to a unique new CSV file
write.csv(CN1_temp, "/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/Seascape_Processed/CN1_temperature.csv")

```

```{r}
#Calculate the salinity variables
salinity_quantile_10 <- quantile(CN1_sal$salinity, 0.1)
salinity_quantile_90 <- quantile(CN1_sal$salinity, 0.9)
Mean_Annual_Salinity_ppt <- mean(CN1_sal$salinity)
Mean_max_Salinity_ppt <- mean(CN1_envryear_sal$max_salinity)
Mean_min_Salinity_ppt <- mean(CN1_envryear_sal$min_salinity)
Salinity_st_dev <- sd(CN1_sal$salinity)
Salinity_n <- nrow(CN1_sal)
Salinity_years <- nrow(CN1_envryear_sal)


#Create a data frame to store the temperature results
CN1_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, Salinity_st_dev, Salinity_n, Salinity_years, salinity_quantile_10, salinity_quantile_90, collection_type)
print(CN1_salinity)

# Write to the combined file with all sites 
write.table(CN1_salinity, "/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/Seascape_Processed/seascape_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE) # The column names should be changed to FALSE after 1st row is added to the data frame

# Write to a unique new CSV file
write.csv(CN1_salinity, "/Users/madelineeppley/GitHub/EasternOysterEnvData/2022_SiteEnvironmentData/Seascape_Processed/CN1_salinity.csv", row.names = FALSE)
```
